---
title: "ClusterSignificance Vignette"
author: "Jesper Robert Gådin and Jason T. Serviss"
date: '`r Sys.Date()`'
output:
    html_document:
        theme: flatly
        toc: yes
vignette: >
    %\VignetteIndexEntry{ClusterSignificance Vignette} 
    %\VignetteEngine{knitr::rmarkdown} 
    %\usepackage[utf8]{inputenc}
    %\SweaveUTF8
---

<style>
    .col2 {
        columns: 2 200px;         /* number of columns and width in pixels*/
        -webkit-columns: 2 200px; /* chrome, safari */
        -moz-columns: 2 200px;    /* firefox */
    }
</style>

```{r style, echo = FALSE, results = 'asis'}
BiocStyle::markdown()
library(ggplot2)
```

# Introduction
The ClusterSignificance package provides tools to assess if clusters, in
e.g. principal component analysis (PCA), have a separation different from
random or permuted data. This is accomplished in a 3 step process *projection*,
*classification*, and *permutation*. To be able to compare cluster
separations, we have to give them a score based on this separation. First, 
all data points in each cluster are projected onto a line (*projection*), after
which the seperation for two groups at a time is scored (*classification*).
Furthermore, to get a p-value for the separation we have to compare the
separation score for our real data to the separation score for permuted data 
(*permutation*).

The package includes 2 different methods for accomplishing the 3 steps
above, Mean line projection (**Mlp**) and Principal curve projection (**Pcp**).
Here we will first discuss the similaraties of the 3 steps independant
of which method is used. This is followed by an example of the Mlp and
Pcp methods and the unique features of each. Furthermore, we provide an
example where ClusterSignificance is used to examine the seperation between
2 clusters downsteam of a PCA. 

```{r echo=TRUE, eval=TRUE, message=FALSE}
library(ClusterSignificance)
```

# Methods
## Projection

The input to the projection methods is a matrix, with data points (or samples)
on the rows, and dimensions (or principal components) in the columns. For
the method to know which rows are in the same group, the groups argument
must be specified. The groups argument is a character vector and 
ClusterSignificance will produce an error if the groups argument contain 
NA's. The projection step utilizes either the Mlp or Pcp method to project 
all data points onto a line, after which they are projected into a single 
dimension.

## Classification

Now that all of the points are in one dimension, it allows us to easily
define a perpendicular line that best separates the two groups. The score
for each possible separation is then calculated based on this formula:

#### $$score = TP + TN - (FP + FN)$$

Where:

* __TP__ = True positives
* __TN__ = True negatives
* __FP__ = False positives
* __FN__ = False negatives

All seperation scores are recorded and the max score is subsequently used
downstream to compare to the max scores from the permutated data.

## Permutation

Permutation is performed by randomly sampling the input matrix and assigning
the data points to the groups. The projection and classification steps
are then run for the sampled matrix. The max seperation score is recorded
and the next permutation is performed. The p-value can subsequently be
calcluated with the following formula:

####$$p.value=\frac{count\ of\ permuted\ scores >= real\ score}{iterations}$$

If none of the permuted scores are greater that the real score the p.value
is instead calculated as:

#### $$p.value < 10^{-log10(iterations)}$$

Ideally, the data is permuted until at least one permuted score is greater
than the real score allowing for the calculation of an exact p-value, although,
with very distinct seperations, this is not always realistic. Despite the
fact that we use a low iteration number in many examples in the vignette,
we recommend 10,000 iterations as a good starting point to examine the
distribution of permuted scores compared to the real score.

# Examples
## Demonstration data

The ClusterSignificance package includes 2 small example data matrices, 
one used for demonstrating the Mlp method and one used for demonstrating 
the Pcp method. The Mlp demo matrix has 2 dimensions whereas, the Pcp demo 
matrix has 4. In addition, the groups variable has been added as the rownames 
of each matrix, with the Mlp demo matrix having 2 groups and the Pcp demo 
matrix having 3 groups. Finally, colnames, indicating the dimension number, 
were also added to each matrix. Neither colnames or rownames are needed for 
the input matrix to ClusterSignificance functions and we have only added 
them here only for clarity.

```{r, demoDataMlp}
data(mlpMatrix)
```

```{r headMat, echo=FALSE, eval=TRUE}
##Mlp
printMat <- mlpMatrix
groups <- rownames(mlpMatrix)
dimnames(printMat) <- list(groups, paste("dim", 1:2, sep=""))
knitr::kable(
    printMat[1:3,], 
    digits = 2, 
    caption = "Head; Mlp matrix.", 
    align=c(rep('l', 2))
)

gTable <- as.data.frame(table(groups))
knitr::kable(
    gTable, 
    digits = 2, 
    align=c(rep('l', 2)), 
    caption = "Rownames table; Mlp matrix."
)
```

```{r, demoDataPcp}
data(pcpMatrix)
```

```{r headPcp, echo=FALSE}
##Pcp
printMat <- pcpMatrix
groups <- rownames(pcpMatrix)
dimnames(printMat) <- list(groups, paste("dim", 1:4, sep=""))
knitr::kable(
    printMat[1:3,], 
    digits = 2, 
    caption = "Head; Pcp matrix.", 
    align=c(rep('l', 2))
)

gTable <- as.data.frame(table(groups))
knitr::kable(
    gTable, 
    digits = 2, 
    align=c(rep('l', 2)), 
    caption = "Rownames table; Pcp matrix."
)
```

## Mlp method
### Projection

The Mlp method is suitable for comparing clusters in a maximum 2 dimensions
when the seperation between a maximum of 2 groups will be evaluated at
a time. Breifly, Mlp functions by drawing a regression line through the
means of the two groups and then projects all points onto that line. To
project the points into one dimension, the points are then rotated onto
the x-axis. A detailed description and graphical representation of the
individual steps can be seen in the example below. It should be noted that,
despite the fact that the Mlp method will work with as low as 2 data points
per group, it may not be advisable to perform an analysis with such a small
amount of data.

We use the demonstration data included with the package to demonstrate 
the Mlp function.

```{r, fig.align='center', fig.width=10, fig.height=8}
## Load demo input.
data(mlpMatrix)

## Create the group variable.
groups <- rownames(mlpMatrix)

## Run Mlp and plot.
prj <- mlp(mlpMatrix, groups)
plot(prj)
```

The plot above shows the 6 steps Mlp takes to project the points into one
dimension.

1. The input matrix.
2. Finding the mean of each group.
3. Regression line and move to origo.
    a. The regression line drawn through the mean of each group.
    b. Adjustment of the line and points so the line passes through origo.
        i. Arrows show the move from the original position to new position.
4. Orthogonal projection onto the regression line.
5. Points shown after projection.
6. The final projection of points into one dimension.

Note that step 4 shows the orthogonal projection onto the regression line
and appears visually to not be perpendicular to the line. This is due to
the automatic scaling of the y-axis which is necessary when large differences
in proportions between PC1 and PC2 would otherwise cause data to disappear
outside of the margins.

### Classification

```{r classifyMlp, message=FALSE, fig.align='center', fig.width=8, fig.height=6}
## Classify and plot.
cl <- classify(prj)
plot(cl)
```

The classification function works in the same manner for both the Mlp and
Pcp methods and is previously described in the [Classification][] section.
The perpendicular lines in the classification plot represent the separation
scores for all possible separations. The highest scoring separation of
the two groups is represented by the highest perpendicular line.

### Permuation

```{r permuteMlp, message=FALSE, fig.align='center', message = FALSE}
## Set the seed and number of iterations.
set.seed(3)
iterations <- 100 

## Permute and plot.
pe <- permute(
    mat=mlpMatrix, 
    iterations=iterations, 
    groups=groups, 
    projmethod="mlp"
)

plot(pe)
```

```{r pValueMlp, echo=FALSE, eval=TRUE, message=FALSE}
## p-value
pvalue(pe)
```

The permutation plot shows a histogram of the max seperation scores from
all permutations with the verrtical red line representing the max score
from the real data. From the plot can see that when comparing the seperation
of grp1 vs. grp2 the permuted data for 100 iterations never scores higher
than the real data, and the separation thus has a significant p-value of
< 0.01. The p-values can be obtained using `pvalue(pe)`.

## Pcp method
### Projection

The Pcp method is suitable in situations where more than 2 dimensions need
to be considered simultaneously and/or more than 2 group seperations should
be analyzed. It is limited by the fact that, the Pcp method will not work
for < 5 data points per group and must have > 5 unique values per dimension.
The Pcp method utilizes the principal.curve method from the princurve package
to fit a principal curve to all data points utilizing the dimensions input
by the user. A principal curve can be described as a "smooth curve that
passes through the middle of the data in an orthogonal sense". All data
points are then projected onto the principal curve and transfered into
one dimension for scoring.

For this example the deomnstration data was simulated so that one cluster
has a visable seperation from the two other clusters, whereas two of the 
clusters are largley overlapping. Therefore, can predict that 
ClusterSignificance will find 2 seperations to be significant where as 
the other will be insignificant.

```{r PcpProjection, fig.align='center', fig.width=10, fig.height=10}
## Load demo input.
data(pcpMatrix)

## Create the group variable.
groups <- rownames(pcpMatrix)

## Run Pcp and plot.
prj <- pcp(pcpMatrix, groups)
plot(prj)
```

The following steps take place during Pcp projection:

1. The input matrix.
2. Drawing the principal curve.
3. Projection of points to the principal curve.
4. The points after projection onto the principal curve.
5. Projection of points into one dimension.
6. The final projection of points.

### Scoring
Due to the fact that the Pcp method can perform several group comparisons
at once, when > 2 groups are used all possible seperations are automatically
analyzed and plotted.

```{r classifyPcp, message=FALSE, fig.width=10, fig.height=7}
cl <- classify(prj)
plot(cl)
```

### Permuation

```{r permutePcp, message=FALSE, fig.width=10, fig.height=7}
#permute matrix
set.seed(3)
iterations <- 100 ## Set the number of iterations.

## Permute and plot.
pe <- permute(
    mat=pcpMatrix, 
    iterations=iterations, 
    groups=groups, 
    projmethod="pcp"
)

plot(pe)
```

```{r pValuePcp, echo=TRUE, eval=TRUE, message=FALSE}
pvalue(pe)
```

The results from the permutation indicate that ClusterSignificance performed
as we would expect. The seperation for grp1 vs grp2 is insignificant whereas
the seperations for both grp1 vs grp3 and grp2 vs grp3 are significant.

# Principal Component Analysis

One probable application for ClusterSignificance would be the analysis
of cluster seperation post-PCA. Here we use the leukemia dataset from the
plsgenomics package to demonstrate how this may be done. The dataset includes
"gene expression data (3051 genes and 38 leukemia samples) from the leukemia
microarray study of Golub et al. (1999)". The tumor samples are from both
acute lymphoblastic leukemia (ALL) and acute myeloid leukemia (AML) patients.
Due to the inherant differences in the transcriptome of these two cancer
types, we hypothesize that they would form seperate culsters post-PCA and
that we can use ClusterSignificance to examine if this seperation is 
significant.


```{r echo=TRUE, eval=TRUE, message=FALSE}
## The leukemia dataset is provided as a list with 3 elements. 
## An explaination of these can be found using the command:
## help(leukemia)

library(plsgenomics, quietly = TRUE)
data(leukemia)

## Extract the expression matrix.
expression <- leukemia$X 

## Run PCA extract the principal components.
pca <- prcomp(expression, scale = TRUE)
mat <- pca$x

## Extract the grouping variable (coded as 1 (ALL) and 2 (AML)).
groups <- ifelse(leukemia$Y == 1, "ALL", "AML")
```

```{r, echo = FALSE}
gTable <- as.data.frame(table(groups))
knitr::kable(gTable, 
    digits = 2, 
    align=c(rep('l', 2)), 
    caption = "Groups argument table."
)
```

```{r ALLprojection, fig.align='center', fig.width=10, fig.height=10}
ob <- pcp(mat, groups, normalize=TRUE)
plot(ob)
```

```{r ALLclassification, fig.align='center', fig.width=8, fig.height=6}
cl <- classify(ob)
plot(cl)
```

```{r ALLpermutation, fig.align='center', message = FALSE}
set.seed(3)
iterations <- 100
pe <- permute(mat=mat, iterations=iterations, groups=groups, projmethod="pcp")
pvalue(pe)
```

As we can see from the resulting p-value, ClusterSignificance is able to
find a significant seperation between ALL and AML clusters, even with the
low iteration number used in the example. 10000 iterations results in p
= 1e-04.

# Advanced usage
## Concatenate permutation results
When running the permutation step of the Pcp method, it is possible that
all iterations are not sucessfully completed, as can be seen with the message
upon permutaion completion. Due to the fact that, the p-value is calculated
not by the number of iterations input by the user but, instead, with the
number of iterations that sucessfully complete, this will not be an issue
for most users. However, if it is the case that a specific number of iterations
is required by the user it is possible to concatenate multiple permutation
runs.  

```{r concatPermuts, message=FALSE, eval=FALSE}
set.seed(3)
iterations <- 10

pe1 <- permute(mat=mat, iterations=iterations, groups=groups, projmethod="pcp")
pe2 <- permute(mat=mat, iterations=iterations, groups=groups, projmethod="pcp")
pe3 <- permute(mat=mat, iterations=iterations, groups=groups, projmethod="pcp")
pvalue(pe1)

pe <- c(pe1, pe2, pe3)
pvalue(pe)
```

## User defined permutatiton matrix
In some cases, e.g. PCA, it may be desireable to define a permutation matrix
rather than allowing ClusterSignificance to do this automatically. For
example, when determining seperation of clusters related to PCA, a more
valid approch may be to perform PCA on a sampled matrix of the origial
values, after which this is provided to ClusterSignificance, rather than
allowing ClusterSignificance to randomly sample the principal components.
This can be accomplished by providing the *user.permutations* argument
to the permute method. 

```{r userDefinedPermats, eval=TRUE}
##define permMatrix function
.pca <- function(y, groups, uniq.groups, mat) {
    pca <- prcomp(
        sapply(1:ncol(mat[groups %in% uniq.groups[, y], ]), 
            function(i)
                mat[groups %in% uniq.groups[, y], i] <- 
                    sample(
                        mat[groups %in% uniq.groups[, y], i], 
                        replace=FALSE
                    )
        ), scale=TRUE
    )
    return(pca$x)
}

permMatrix <- function(iterations, groups, mat) {
    uniq.groups <- combn(unique(groups), 2)
    permats <- lapply(1:ncol(uniq.groups), 
        function(y)
            lapply(1:iterations, 
                function(x)
                    .pca(y, groups, uniq.groups, mat)
            )
    )
    return(permats)
}

set.seed(3)
points.amount <- 20
groupNames = c("grp1", "grp2", "grp3")
PCs = 4

## Create the input matrix.
mat <- matrix(c(
    sapply(1:PCs, function(xi)
        c(
            sample(seq(0.0, 1.0, 0.001), replace=FALSE, size=points.amount),
            sample(seq(0.0, 1.0, 0.001), replace=FALSE, size=points.amount),
            sample(seq(0.0, 1.0, 0.001), replace=FALSE, size=points.amount)
        )
    )), ncol=PCs
) 

## Create the group variable.
groups <- c(
    sapply(1:length(groupNames), function(ix, groupNames)
        rep(groupNames[ix], nrow(mat)/length(groupNames)), 
        groupNames = groupNames
    )
)

iterations = 100
permats <- permMatrix(iterations, groups, mat)
```

This results in a list with one element per comparison. Each of these elements
holds another list with the number of elements = number of iterations and
each element holding 1 permutation matrix. We can then provide this permutation
matrix to ClusterSignificance in the following manner:

```{r runUserDefinedMatrix, eval=FALSE}
pe <- permute(
    mat = mat, 
    iterations=100, 
    groups=groups, 
    projmethod="pcp",
    userPermats=permats
)
```

# Conclusion

This package can estimate a p-value for a separation projected on a line.
Typically the package might be used post-PCA-analysis, but also clusters
coming from other sources can be tested.

# Session info

```{r echo=TRUE, eval=TRUE}
sessionInfo()
```

